Fast Random Integer Generation in an Interval

### 2024-02-15 Presentation/unbiased.tex
finished presentation v1

\begin{frame}{The OpenBSD Algorithm}
    Algorithm: \begin{itemize}
        \item Generate a uniform random number $x \in [0,2^W)$ until $x \geq \RNW$
        \item Return $x \Mod n$
    \end{itemize} 
    \pause 
    \vspace*{0.5cm}
    \begin{block}{Efficiency}<+->
        \onslide<+->{We require $2$ integer division operations:} \begin{itemize}[<+->]
            \item one for computing $\RNW$
            \item and one for computing $x \Mod n$.
        \end{itemize}
    \end{block}
\end{frame}

\subsection{The Java Algorithm}\label{sec:2.2}
\begin{frame}{The Java Algorithm}
    \pause 
    \begin{itemize}
        \item<1-> Consider $x - \left(x \Mod n\right)$ for $x \in [0,2^W)$: \only<3>{\begin{align*}\everymath{} % Allows specific coloring
            \overbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
        \end{align*}} \only<4>{\begin{align*}\everymath{} % Allows specific coloring
            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
        \end{align*}} \only<5>{\begin{align*}\everymath{} % Allows specific coloring
            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
        \end{align*}} \only<6>{\begin{align*}\everymath{} % Allows specific coloring
            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\substack{\text{mapped to} \\ 2^W - n - \RNW}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
        \end{align*}} \only<7->{\begin{align*}\everymath{} % Allows specific coloring
            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\substack{\text{mapped to} \\ 2^W - n - \RNW}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\underbrace{\overbrace{\BB{0},1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}_{\substack{\text{mapped to} \\ 2^W - \RNW}}}
        \end{align*}}
        \item<8-> Map every number to the next-smallest multiple of $n$
        \item<9-> Only numbers in \ER{leftover} interval mapped to $\ER{2^W - \RNW} > 2^W - n$
        \item<10-> Algorithm: \begin{itemize}
            \item[(1)]<11-> Draw $x \in [0,2^W)$ and compute $r = x \Mod n$
            \item[(2)]<12-> Return $r$ if $x - r > 2^W - n$ else goto \textbf{\GB{(1)}}
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{The Java Algorithm}
    Algorithm: \begin{itemize}
        \item[(1)] Draw $x \in [0,2^W)$ and compute $r = x \Mod n$
        \item[(2)] Return $r$ if $x - r > 2^W - n$ else goto \textbf{\GB{(1)}}
    \end{itemize}
    \pause 
    \smallskip
    \begin{block}{Efficiency}<+->
        \begin{itemize}[<+->]
            \item At least one integer division operation
            \item Number of integer divisions operations equal to number of \GB{rounds}
            \item Return number in \GB{round} if $x < 2^W - \RNW$
            \item Happens with probability $\frac{2^W - \RNW}{2^W} > \frac{1}{2}$
            \item Expected number of integer division operations is $\frac{2^W}{2^W - \RNW} < 2$
        \end{itemize}
    \end{block}
\end{frame}

\subsection{The Bitmask Algorithm}\label{sec:2.3}
\begin{frame}{The Bitmask Algorithm}
    \pause 
    \begin{itemize}[<+->]
        \item Consider the \GB{binary} representation of $n$: \onslide<+->{\begin{align*}\everymath{}
            n \quad \xrightarrow{\text{binary}} \qquad \underbrace{
                \overbrace{0}^{
                    \hidewidth 2^{W - 1} \hidewidth 
                },\ldots,0
            }_{\text{only $0$'s}},\ER{\overbrace{1}^{
                 2^{\lfloor\log_2n\rfloor}
            }},\underbrace{
                \nicefrac{0}{1},\ldots,\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^1 },\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^0 }
            }_{\text{series of $0$'s and $1$'s}}
        \end{align*}} 
        \item Every number $x \leq n$ only needs the last $\lfloor\log_2n\rfloor + 1$ bits
        \item Get these bits with a bitwise-\textsc{And} with \begin{align*}\everymath{}
            2^{\lfloor\log_2n\rfloor + 1} - 1 \quad \xrightarrow{\text{binary}} \qquad \underbrace{
                \overbrace{0}^{
                    \hidewidth 2^{W - 1} \hidewidth 
                },\ldots,0
            }_{\text{only $0$'s}},\underbrace{\ER{\overbrace{1}^{
                 2^{\lfloor\log_2n\rfloor}
            }},1,\ldots,\overbrace{
                    1
                }^{ 2^1 },\overbrace{
                    1
                }^{ 2^0 }
            }_{\text{only $1$'s}}
        \end{align*}
    \end{itemize}
\end{frame}

\begin{frame}{The Bitmask Algorithm}
    \pause 
    \begin{itemize}[<+->]
        \item How can we compute $2^{\lfloor\log_2n\rfloor + 1}$? 
        \item \GB{Count} the number $\ell$ of \GB{leading $0$'s}! \onslide<+->{\begin{align*}\everymath{}
            n \quad \xrightarrow{\text{binary}} \qquad \underbrace{
                \overbrace{0}^{
                    \hidewidth 2^{W - 1} \hidewidth 
                },\ldots,0
            }_{\text{$\ell$ $0$'s}},\ER{\overbrace{1}^{
                 2^{\lfloor\log_2n\rfloor}
            }},\underbrace{
                \nicefrac{0}{1},\ldots,\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^1 },\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^0 }
            }_{\text{series of $0$'s and $1$'s}}
        \end{align*}} 
        \item $\ER{\lfloor\log_2n\rfloor} = W - \ell - 1$ \onslide<+->{$\qquad \longrightarrow \qquad 2^{\lfloor\log_2n\rfloor + 1} = 1 \ll \left(W - \ell\right)$}
        \item Algorithm: \begin{itemize}[<+->]
            \item[(1)] Compute $\ell$ and $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
            \item[(2)] Draw $x \in [0,2^W)$ and compute $b = x \And \mathcal{M}$
            \item[(3)] Return $b$ if $b < n$ else goto \textbf{\GB{(2)}}
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{The Bitmask Algorithm}
    Algorithm: \begin{itemize}
        \item[(1)] Compute $\ell$ and $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
        \item[(2)] Draw $x \in [0,2^W)$ and compute $b = x \And \mathcal{M}$
        \item[(3)] Return $b$ if $b < n$ else goto \textbf{\GB{(2)}}
    \end{itemize}
    \pause 
    \smallskip
    \begin{block}{Efficiency}<+->
        \begin{itemize}[<+->]
            \item $b$ at most $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1 < 2n$ \onslide<+->{$\quad \longrightarrow \quad$ success probability at least $\approx \frac{1}{2}$}
            \item At most $\approx 2$ rounds in expectation
            \item \GB{No} integer division at all
            \item Computation of \GB{leading $0$'s} requires \GB{\texttt{clz}} instruction/algorithm 
            \item Roughly as expensive as a \GB{\texttt{div}} instruction 
        \end{itemize}
    \end{block}
\end{frame}

%%% Array Drawing with Tikz instead of Equation + FastDiceRoller
\iffalse
\onslide<+->{\begin{center}
    \begin{tikzpicture}[scale = 0.7]
        \draw[draw = goetheblau] (0.0,0.0) rectangle ++(5.0,1.0);
        \draw[draw = goetheblau,overbrace style] (0.0,1.0) -- (5.0,1.0) node [overbrace text style] {\GB{$n$ values}};
        \draw[draw = goetheblau] (5.0,0.0) rectangle ++(5.0,1.0);
        \draw[draw = goetheblau,overbrace style] (5.0,1.0) -- (10.0,1.0) node [overbrace text style] {\GB{$n$ values}};
        \node at (10.5,0.5) {$\cdots$};
        \draw[draw = goetheblau] (11.0,0.0) rectangle ++(5.0,1.0);
        \draw[draw = goetheblau,overbrace style] (11.0,1.0) -- (16.0,1.0) node [overbrace text style] {\GB{$n$ values}};
        \draw[draw = emorot] (16.0,0.0) rectangle ++(3.0,1.0);
        \draw[draw = emorot,overbrace style] (16.0,1.0) -- (19.0,1.0) node [overbrace text style] {$\text{\ER{\RNW values}}$};
    \end{tikzpicture}
\end{center}}

\subsection{The Fast-Dice-Roller Algorithm}
\begin{frame}{The Fast-Dice-Roller Algorithm}
    \pause 
    \begin{itemize}[<+->]
        \item Build up $x$ bit-by-bit using uniform random bits \GB{\texttt{flip()}}
        \item Keep track of upper bound $\mathcal{B}$ for number \onslide<+->{  $\longrightarrow \quad x \in [0,\mathcal{B})$}
        \item Repeat until $\mathcal{B} \geq n$ \begin{itemize}[<+->]
            \item if $x < n$, return $x$
            \item else decrease $x$ and $\mathcal{B}$ by $n$ (\ER{rejection})
        \end{itemize}
    \end{itemize}

\end{frame}
\fi

>>>>

\begin{frame}{The OpenBSD Algorithm - Efficiency}
    Algorithm: \begin{itemize}
        \item Generate a uniform random number $x \in [0,2^W)$ until $x \geq \RNW$
        \item Return $x \Mod n$
    \end{itemize} 
    \pause 
    \vspace*{0.5cm}
    \begin{block}{Efficiency}<+->
        \onslide<+->{We require $2$ integer division operations:} \begin{itemize}[<+->]
            \item one for computing $\RNW$
            \item and one for computing $x \Mod n$.
        \end{itemize}
    \end{block}
\end{frame}

\subsection{The Java Algorithm}\label{sec:2.2}
\begin{frame}{The Java Algorithm}
    \pause 
    \begin{itemize}
        \item<1-> Consider $x - \left(x \Mod n\right)$ for $x \in [0,2^W)$: \only<3>{\begin{align*}\everymath{} % Allows specific coloring
            \overbrace{\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
        \end{align*}} \only<4>{\begin{align*}\everymath{} % Allows specific coloring
            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
        \end{align*}} \only<5>{\begin{align*}\everymath{} % Allows specific coloring
            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\overbrace{0,1,\ldots,n - 1}^{\text{$n$ values}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
        \end{align*}} \only<6>{\begin{align*}\everymath{} % Allows specific coloring
            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\substack{\text{mapped to} \\ 2^W - n - \RNW}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\overbrace{0,1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}
        \end{align*}} \only<7->{\begin{align*}\everymath{} % Allows specific coloring
            \overbrace{\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $0$}},\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\text{mapped to $n$}},\ldots,\underbrace{\overbrace{\BB{0},1,\ldots,n - 1}^{\text{$n$ values}}}_{\substack{\text{mapped to} \\ 2^W - n - \RNW}}}^{\text{$\left(2^W \div n\right) \cdot n$ values}},\ER{\underbrace{\overbrace{\BB{0},1,\ldots,\RNW - 1}^{\text{$\RNW$ values}}}_{\substack{\text{mapped to} \\ 2^W - \RNW}}}
        \end{align*}}
        \item<8-> Map every number to the next-smallest multiple of $n$
        \item<9-> Only numbers in \ER{leftover} interval mapped to $\ER{2^W - \RNW} > 2^W - n$
        \item<10-> Algorithm: \begin{itemize}
            \item[(1)]<11-> Draw $x \in [0,2^W)$ and compute $r = x \Mod n$
            \item[(2)]<12-> Return $r$ if $x - r \leq 2^W - n$ else goto \textbf{\GB{(1)}}
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{The Java Algorithm - Efficiency}
    Algorithm: \begin{itemize}
        \item[(1)] Draw $x \in [0,2^W)$ and compute $r = x \Mod n$
        \item[(2)] Return $r$ if $x - r \leq 2^W - n$ else goto \textbf{\GB{(1)}}
    \end{itemize}
    \pause 
    \smallskip
    \begin{block}{Efficiency}<+->
        \begin{itemize}[<+->]
            \item At least one integer division operation
            \item Number of integer divisions operations equal to number of \GB{rounds}
            \item Return number in \GB{round} if $x < 2^W - \RNW$
            \item Happens with probability $\frac{2^W - \RNW}{2^W} > \frac{1}{2}$
            \item Expected number of integer division operations is $\frac{2^W}{2^W - \RNW} < 2$
        \end{itemize}
    \end{block}
\end{frame}

\subsection{The Bitmask Algorithm}\label{sec:2.3}
\begin{frame}{The Bitmask Algorithm - Representation}
    \pause 
    \begin{itemize}[<+->]
        \item Consider the \GB{binary} representation of $n$: \onslide<+->{\begin{align*}\everymath{}
            n \quad \xrightarrow{\text{binary}} \qquad \underbrace{
                \overbrace{0}^{
                    \hidewidth 2^{W - 1} \hidewidth 
                },\ldots,0
            }_{\text{only $0$'s}},\ER{\overbrace{1}^{
                 2^{\lfloor\log_2n\rfloor}
            }},\underbrace{
                \nicefrac{0}{1},\ldots,\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^1 },\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^0 }
            }_{\text{series of $0$'s and $1$'s}}
        \end{align*}} 
        \item Every number $x \leq n$ only needs the last $\lfloor\log_2n\rfloor + 1$ bits
        \item Get these bits with a bitwise-\textsc{And} with \begin{align*}\everymath{}
            2^{\lfloor\log_2n\rfloor + 1} - 1 \quad \xrightarrow{\text{binary}} \qquad \underbrace{
                \overbrace{0}^{
                    \hidewidth 2^{W - 1} \hidewidth 
                },\ldots,0
            }_{\text{only $0$'s}},\underbrace{\ER{\overbrace{1}^{
                 2^{\lfloor\log_2n\rfloor}
            }},1,\ldots,\overbrace{
                    1
                }^{ 2^1 },\overbrace{
                    1
                }^{ 2^0 }
            }_{\text{only $1$'s}}
        \end{align*}
    \end{itemize}
\end{frame}

\begin{frame}{The Bitmask Algorithm - Mask}
    \pause 
    \begin{itemize}[<+->]
        \item How can we compute $2^{\lfloor\log_2n\rfloor + 1}$? 
        \item \GB{Count} the number $\ell$ of \GB{leading $0$'s} in $n$! \onslide<+->{\begin{align*}\everymath{}
            n \quad \xrightarrow{\text{binary}} \qquad \underbrace{
                \overbrace{0}^{
                    \hidewidth 2^{W - 1} \hidewidth 
                },\ldots,0
            }_{\text{$\ell$ $0$'s}},\ER{\overbrace{1}^{
                 2^{\lfloor\log_2n\rfloor}
            }},\underbrace{
                \nicefrac{0}{1},\ldots,\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^1 },\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^0 }
            }_{\text{series of $0$'s and $1$'s}}
        \end{align*}} 
        \item $\ER{\lfloor\log_2n\rfloor} = W - \ell - 1$ \onslide<+->{$\qquad \longrightarrow \qquad 2^{\lfloor\log_2n\rfloor + 1} = 1 \ll \left(W - \ell\right)$}
        \item Algorithm: \begin{itemize}[<+->]
            \item[(1)] Compute $\ell$ and $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
            \item[(2)] Draw $x \in [0,2^W)$ and compute $b = x \And \mathcal{M}$
            \item[(3)] Return $b$ if $b < n$ else goto \textbf{\GB{(2)}}
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{The Bitmask Algorithm - Efficiency}
    Algorithm: \begin{itemize}
        \item[(1)] Compute $\ell$ and $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
        \item[(2)] Draw $x \in [0,2^W)$ and compute $b = x \And \mathcal{M}$
        \item[(3)] Return $b$ if $b < n$ else goto \textbf{\GB{(2)}}
    \end{itemize}
    \pause 
    \smallskip
    \begin{block}{Efficiency}<+->
        \begin{itemize}[<+->]
            \item $b$ at most $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1 < 2n$ \onslide<+->{$\quad \longrightarrow \quad$ success probability at least $\approx \frac{1}{2}$}
            \item At most $\approx 2$ rounds in expectation
            \item \GB{No} integer division at all
            \item Computation of \GB{leading $0$'s} requires \GB{\texttt{clz}} instruction/algorithm 
            \item Roughly as expensive as a \GB{\texttt{div}} instruction 
        \end{itemize}
    \end{block}
\end{frame}
